OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-04-06 paper/paper.Rmd
Restructure ref-db section

@@

- TODO: start paragraph with question, end with why we should care. transitions to move the story along.
- start:"In order to compare Optifit.."
- end: "optifit performs better"
- AND, BUT, THEREFORE, structure
- order of importance: mcc, mapping, sec (1 paragraph each, in this order)

While _de novo_ clustering produces high quality OTUs, researchers may prefer to
perform reference clustering to a public database because reference-based
methods produce stable OTUs and are generally faster than _de novo_ methods.
To test how OptiFit performs for this purpose, we fit each dataset to three
databases of reference OTUs; the Greengenes database, the SILVA non-redundant
database, and the Ribosomal Database Project (RDP) [@desantis_greengenes_2006;
@quast_silva_2013; @cole_ribosomal_2014].
Reference OTUs for each database were created by performing _de novo_ clustering
with OptiClust at a distance threshold of 3%.
To compare to existing software, we also clustered each dataset into _de novo_
OTUs using OptiClust and VSEARCH, and with VSEARCH in reference-based mode
against the Greengenes database of OTUs clustered at 97% similarity.

In closed reference mode against databases, OptiFit produced lower quality OTUs
than OptiClust by `r mcc_closed_fit_db_vs_clust` on average.
Fitting sequences to Greengenes and SILVA in closed reference mode performed
similarly with median MCC scores of `r closed_fit_gg_mcc` and 
`r closed_fit_silva_mcc` respectively, while when fitting to RDP the median MCC 
dropped to `r closed_fit_rdp_mcc`.
An average of only `r frac_fit_db`% of query sequences were fit to reference
database OTUs in closed reference mode across all dataset/database combinations.
VSEARCH was able to map `r frac_vsearch_vs_fit`% more query sequences than
OptiFit to the Greengenes reference database.
In terms of run time, closed reference OptiFit outperformed OptiClust by 
`r sec_closed_fit_db_vs_clust`% and VSEARCH by `r sec_closed_fit_db_vs_vsearch`%.

since closed ref threw out a lot of stuff, we performed open ref.

In open reference mode, fitting the datasets to reference OTUs with OptiFit
produced OTUs of similar quality (`r mcc_open_fit_db_vs_clust`% difference in median
MCC) as clustering the datasets _de novo_ with OptiClust  across all datasets
and reference databases.
OptiFit produced higher quality OTUs than VSEARCH when open reference clustering
against the Greengenes database, with median MCC scores of `r open_fit_gg_mcc`
and `r open_vsearch_mcc` (respectively).
OptiFit ran faster than VSEARCH by `r sec_vsearch_vs_open_fit_db`%  in open
reference mode, but slower than OptiClust by `r sec_opticlust_vs_open_fit_db`%.
_De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
higher quality OTUs than VSEARCH, but performed `r sec_opticlust_vs_vsearch`% 
slower than VSEARCH.

>>>>

- **start paragraph with question, end with why we should care. transitions to move the story along.**
- **AND, BUT, THEREFORE, structure**

@@

While _de novo_ clustering produces high quality OTUs, researchers may prefer to
perform reference clustering to a public database because reference-based
methods produce stable OTUs and are generally faster than _de novo_ methods.
In closed reference mode, sequences that cannot be assigned to reference OTUs
are thrown out, so that the final clustering contains only OTUs that exist in
the reference.
To test how OptiFit performs for this purpose, we fit each dataset to three
databases of reference OTUs; the Greengenes database, the SILVA non-redundant
database, and the Ribosomal Database Project (RDP) [@desantis_greengenes_2006;
@quast_silva_2013; @cole_ribosomal_2014].
Reference OTUs for each database were created by performing _de novo_ clustering
with OptiClust at a distance threshold of 3%.
Fitting sequences to Greengenes and SILVA in closed reference mode performed
similarly, with median MCC scores of `r closed_fit_gg_mcc` and 
`r closed_fit_silva_mcc` respectively, while when fitting to RDP the median MCC 
dropped to `r closed_fit_rdp_mcc`.
For comparison to the state of the art, clustering datasets _de novo_ with
OptiClust produced an average MCC score of `r opticlust_mcc`.
This gap in OTU quality mostly disappears when clustering in open reference
mode, which produced median MCCs of `r open_fit_gg_mcc` with greengenes, 
`r open_fit_silva_mcc` with Silva, and `r open_fit_rdp_mcc` with RDP.
Thus, open reference OptiFit produces OTUs of very similar quality as _de novo_
clustering, and closed reference OptiFit follows closely behind as long as a
suitable reference database is chosen.

Since closed reference clustering throws out query sequences that cannot be
assigned to reference OTUs, an additional measure of clustering performance to
consider is the fraction of query sequences that were able to be assigned.
On average, more sequences were assigned with Greengenes as the reference 
(`r frac_fit_gg`%) than with Silva (`r frac_fit_silva`%) or RDP (`r frac_fit_rdp`%).
This mirrors the result reported above that Greengenes produced better OTUs in
terms of MCC score than either Silva or RDP.
Note that _De novo_ and open reference clustering methods always assign 100% of
sequences to OTUs.
The database chosen affects the final OTU assignments considerably in terms of 
both MCC score and fraction of query sequences fit.

Despite the drawbacks, closed reference methods have been used when fast
execution speed is required such as when using very large datasets.
To compare performance in terms of speed, we repeated each OptiFit and OptiClust
run 100 times and measured the execution time.
Closed reference OptiFit outperformed OptiClust by 
`r sec_closed_fit_db_vs_clust`% and VSEARCH by `r sec_closed_fit_db_vs_vsearch`%.
In open reference mode, OptiFit ran faster than VSEARCH by 
`r sec_vsearch_vs_open_fit_db`%, but slower than OptiClust by 
`r sec_opticlust_vs_open_fit_db`%.

To compare to the reference clustering method used by QIIME2, we clustered each
dataset with VSEARCH against the Greengenes database of OTUs previously 
clustered at 97% sequence similarity.
**TODO note vsearch otu definition vs. ours.**
_De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
higher quality OTUs than VSEARCH, but performed `r sec_opticlust_vs_vsearch`% 
slower than VSEARCH.
VSEARCH was able to map `r frac_vsearch_vs_fit`% more query sequences than
OptiFit to the Greengenes reference database in closed reference mode.
OptiFit produced higher quality OTUs than VSEARCH when open reference clustering
against the Greengenes database, with median MCC scores of `r open_fit_gg_mcc`
and `r open_vsearch_mcc` (respectively).
